Advice on developing ordinal outcomes and endpoints for clinical trials

### 2025-04-23 paper.tex
Lots of text + pmax efficiency figure

\title{Guidance on choosing ordinal outcomes and endpoints for clinical trials}
\author{Leon Di Stefano, Katherine Lee, etc\ldots}
\date{\today}

@@

\tableofcontents

@@

We want to distinguish \emph{choosing} an outcome/effect
measure from \emph{validating} the outcome/effect measure using
external data. This paper is about choosing the outcome and effect
measure. In practice choosing an outcome and effect measure and
validating that outcome and effect measure in pre-trial data will be
an iterative process.

@@

We want effect measures that
\begin{enumerate}
  \item capture decision-relevant differences in patient benefit and harm
  \item are easy to interpret and to translate into practical recommendations
\end{enumerate}

@@

\subsection{Tie-breaking can be used to fine-grain ordinal outcomes}

Two kinds: crossing, and tie breaking within a level (overriding).

@@

\subsection{Ordinal outcomes can incorporate longitudinal or
time-to-event information}

Distinguish between ``state'' and ``trajectory'' variables.

Trajectory variables, including times-to-event, summarize over a timecourse.

These can be incorporated into an ordinal outcome. For example ``worst pSOFA
while hospitalized until day 28''.

Arguably most existing time-to-event analyses are in fact analyses of an
ordinal outcome. The Cox model treats time-to-event as an ordinal outcome:
because (1) the baseline hazard is not modelled, (2) the model itself is
invariant to monotone rescalings of the time axis, and (3) the (partial)
likelihood (ignoring ties) depends only on the ranks of the data (i.e., is a
marginal likelihood).

@@

\subsection{Model-based versus nonparametric effect measures}

@@

Nonparametric or model-free measures: probabilistic index, win ratio, win odds,
...

Both of these can be considered ``estimands'', insofar as we consider.

Nonparametric effect measures may appear to be assumption-free. However
\begin{itemize}
  \item Assumptions don't need to hold exactly for the model-based
    effect measures to make sense
  \item The model-free methods also may not make sense, if for
    example there are nonmonotone effects across cutpoints
  \item Model-based methods have additional advantages, for example
    the possibility of adjustment for covariates.

    When additivity approximately holds, this yields a ``best of both
    worlds'' effect measure that is
    personalized effect measure that is yet reportable as a single
    number. Risk ratio for vaccine efficacy (being approximately
    constant as a function of baseline risk/age) as an example.
\end{itemize}

@@

\subsection{Generalized pairwise comparisons (``win'' methods)
  are implicitly about constructing
ordinal outcomes, in a way that is compatible with right-censoring}

>>>>

\title{Advice on developing ordinal outcomes and endpoints for clinical trials}
\author{Leon Di Stefano, Katherine Lee, etc\ldots}
\date{\today}

@@

\tableofcontents
\newpage

@@

We want to distinguish \emph{choosing} an outcome/effect measure from
\emph{validating} the outcome/effect measure using external data. This paper is
about the first task, choosing (or constructing) the outcome and effect
measure. In practice choosing an outcome and effect measure and validating that
outcome and effect measure in pre-trial data will be an iterative process; our
focus is on the ``positive'' side of the process, namely coming up
with, and tweaking, proposed outcomes and effect measures.

@@

We want effect measures that
\begin{enumerate}
  \item capture decision-relevant differences in distributions of
    patient benefit and harm
  \item are easy to interpret and to translate into practical recommendations
\end{enumerate}

@@

\subsection{Tie-breaking can be used to combine and fine-grain ordinal outcomes}

Two basic operations:
\begin{enumerate}
  \item Complete tie-breaking of $A$ by
    $B$ (lexicographic order)
  \item Tie-breaking of $A$ by $B$ \emph{within a
    specific level $a$ of $A$} (equivalently, ``overriding'' $B$ by $A$) is the
    combination of these.
\end{enumerate}

Tie breaking within a level (2) is a combination of complete
tie-breaking (1) and merging.

\subsubsection{Incorporating longitudinal or time-to-event information}

Distinguish between ``state'' and ``trajectory'' variables. State
variables summarize patient outcomes at a particular moment in time.
Trajectory variables, including times-to-event, summarize patient
state over a timecourse.

Trajectory variables can be incorporated into an ordinal outcome. For example
(from PASSPORT) ``worst pSOFA while hospitalized until day 28''.

Incorporating trajectory variables into an ordinal outcome may seem unnatural.
But arguable most existing time-to-event analyses are in fact analyses of an
ordinal outcome. The Cox model for time-to-event data treats time as ordinal:
because (1) the baseline hazard is not modelled, (2) the model itself is
invariant to monotone rescalings of the time axis, and (3) the (partial)
likelihood (ignoring ties) depends only on the ranks of the data (i.e., is a
marginal likelihood). The Cox model is essentially an ordinal
stopping ratio model with complementary log-log link. (Perhaps a
difference is censoring.)

\subsection{Coarsening (composite outcomes) can be used to avoid
directly comparing levels}

This comes from the experience of presenting the PASSPORT outcome:
clinicians thought that including, for example, "digit amputation
before day 28" and "ECMO before day 28" within the same level of an
ordinal outcome meant assuming that the two outcomes have the
\emph{same} utility. In fact it represents \emph{refusing to compare}
their utilities.

@@

Theory of \citep{} says that the relative efficiency (variance) of an
unadjusted treatment comparison using an ordinal versus continuous outcome in
the proportional odds model is given by $1 - \sum \overline p_c^3$ where
$\overline p_c$ is the marginal probability (averaged over control and
experimental groups) of outcome category $c$.

Efficiency is maximized at $1 - 1/C^2$ where $C$ is the total number of
categories when the probabilities are all equal: $\overline p_c = 1/C$.

We have simulation results and some math to show that the value of $\sum_c
\overline p_c^3$ is \emph{essentially a function of the largest category
probability} $\overline{p_{\max}} = \max \overline{p_c}$. See \ref{fig:p_max}
for an illustration.

\begin{figure}
  \includegraphics[width=6in]{p_max_controls_efficiency.png}
  \caption{Bounds on the relative efficiency of an ordinal outcome
    compared to its continuous counterpart as a function of the maximum
  category probability.}
  \label{fig:p_max}
\end{figure}

Thus simple advice is: for statistical efficiency, try to make the proportion
of patients in the largest category small.

@@

The basic problem: we want to compare two distributions (control and
experimental groups) over ordinal outcomes with a single number.

In general this represents a loss of information. Let's say the
outcome has $L$ levels. Then the distributions in each
arm is specified by $L-1$ numbers (for the first $L-1$ probabilities,
  or cumulative logits, or stopping ratio logits--reflecting the
sum-to-1 constraint on probability distributions) and so the
\emph{difference} between the distributions is also specified by
$L-1$ numbers (for example the differences between the cumulative
logits, or logit stopping probabilities).

An effect measure thus represents a way to collapse these $L-1$
numbers into one.

\subsection{Model-based effect measures}

@@

What they have in common: the effect measure is a byproduct of fitting a model
to the data. The model typically assumes that a certain pattern holds in the
outcome distributions across groups so that the differences in distributions
can be described by a single number (with more than two groups, $G-1$ numbers).

\subsection{Model-free effect measures}

Nonparametric or model-free measures: probabilistic index, win ratio, win odds,
etc.

\subsubsection{Connection with win win/generalized pairwise comparison
methods}

Win/generalized pairwise comparison methods are implicitly about (1)
constructing ordinal outcomes through tie-breaking (2) in a way that is
compatible with right-censoring (3) estimating nonparametric effect
measures (4) defining comparisons of two distributions by averaging
over draws from the distributions as if they were independent
(pairwise comparisons).

\subsection{Comparing model-based and nonparametric effect measures}

Both of these can be considered ``estimands'', insofar as we consider
the model-based effect measures to be defined by the estimate from
the best-fitting model in the population.

Nonparametric effect measures may appear to be assumption-free. However

\begin{itemize}
  \item Assumptions don't need to hold exactly for the model-based
    effect measures to make sense
  \item The model-free methods also may not make sense, if for
    example there are nonmonotone effects across cutpoints
  \item Model-free methods don't define a common scale for comparing
    more than two interventions. For example, model-free effect
    measures can be nontransitive
  \item Model-based methods have additional advantages, for example
    the possibility of adjustment for covariates.

    When additivity approximately holds, this yields a ``best of both
    worlds'' effect measure that is
    personalized effect measure that is yet reportable as a single
    number.
\end{itemize}

@@

We should distinguish between effect measures used for
\emph{analysis} and for \emph{reporting}. It may make sense to report
the probabilistic index even if the analysis is undertaken using a
proportional odds model.

\section{Conclusions}
